---
title: "Lecture 21: various statistical dangers"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(openintro)
```


<div id="boxedtext">
**Learning Objectives**: 

- Understand the problem with running many simultaneous statistical tests, as we do in multiple linear regression. 
- Understand the most basic solution to the multiple testing problem: Bonferroni correction.
- Understand how model selection (automated or not!) invalidates our subsequent inference.

</div>


# Generating data from a totally null population

First, I write a function that will give me a totally null random dataset. The dataset will have $n$ observations, $p$ identical predictor variables $X_1, \ldots, X_p$, and a single response variable $y$. 

```{r, echo=T, eval=T}
generate.null <- function(n,p) {
  X <- matrix(data=rnorm(n*p, mean=0, sd=1), nrow=n)
  y <- rnorm(n, mean=0, sd=3)
  dat <- data.frame(cbind(y,X))
  names(dat)[2:(p+1)] <- paste0("x", 1:p)
  return(dat)
}
```

This is the first time that we have written our own function in this class. You can play around with how this works below. 

```{r, eval=F}
generate.null(10,3)
generate.null(5,8)
```








## Train/test splits

*** 
## Acknowledgements

The formatting of this tutorial was adopted from an OpenIntro lab. <div id="license">
This is a product of OpenIntro that is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
</div>

