---
title: "Stat 202: Exploratory Data Analysis in R"
author: "Part 2"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(openintro)
```


<div id="boxedtext">

**Learning Objectives**

- Continue practicing with the RStudio interface.
- Re-load and re-clean the data from Tuesday. 
- Create numerical and graphical summaries of variables. 
- Start to explore relationships between variables.
</div>

# Getting started

I am hoping that everyone worked through at least part of **tutorial 1** (https://anna-neufeld.github.io/tutorials/202_fall26/day1.html), and so today's tutorial has less introductory material.

Today, we are going to revisit the parts of tutorial 1 that we skipped or did not get to, and then we will move into new material.

# Loading our data and our packages

Let's load the same dataset that we used on Tuesday. As a reminder, I posted this dataset on my github website for maximum convenience, and you can load it with the following code. If you are working in an RMarkdown file, make sure to load the tidyverse in this file as well. We will need this *package* for data wrangling and plotting!

```{r}
library(tidyverse)
form_data <- read.csv("https://anna-neufeld.github.io/tutorials/202_fall26/fall_response_anon.csv")
```

Last class, we *cleaned* the ``pets`` variable in this dataset. But the version that I just loaded in from the URL above is not clean! I need to rerun the code from tutorial 1 to convert ``pets`` to a numerical variable.

```{r}
form_data$pets[15] <- 1
form_data$pets[36] <- 2
form_data <- form_data %>% mutate(pets = as.numeric(pets))
```

If you are not sure why we did the steps above, please see *tutorial 1*. For now, we will move on and feel confident that we can treat ``pets`` as a numerical variable.

Here are our goals for today. 

<div id="boxedtext">

**Goals:**

- How do we numerically and visually summarize one numerical variable?
- How do we numerically and visually summarize one categorical variable?
- How do we explore relationships between variables? 
</div>

# Some quick review from last class

### Extracting columns with ``select()``

There are generally many ways to do a single thing in R. You all saw last class that the following four lines of code all do essentially the same thing: they extract the ``pets`` variable from the ``form_data`` dataset for us. Note that the last two lines will not work unless you have already run ``library(tidyverse)`` (and included this line in your .Rmd document, if you hope to knit this document later). 

```{r}
form_data$pets
form_data[,2]
form_data %>% pull(pets)
form_data %>% select(pets)
```

**Bonus:** Can you figure out which command above is not like the others? What makes it different?

In general, we are going to prefer ``tidyverse`` methods in this class. So, when you want to extract a column from a dataset, it's nice to use ``select()`` or ``pull()``. We like ``tidyverse`` methods because they can easily be combined with one another, as we will see below. 

### Extracting rows with ``filter()``

When we want to select a subset of *observations* (rows) from a dataset, we use ``filter()``. The following code displays only the data for students who play a sport at Williams. 

```{r, eval=F}
form_data %>% filter(sports=="Yes")
```

Suppose that we want to save the data subset above to a new dataset called ``athletes``. We can certainly do that!

```{r, eval=F}
athletes <- form_data %>% filter(sports=="Yes")
```

One reason that the ``tidyverse()`` set of functions is so nice is because it becomes quite easy to combine functions together. What does the following code do?

```{r, eval=F}
form_data %>% filter(sports=="Yes") %>% select(pets)
```

And, as a harder question, why doesn't the following code work? 

```{r, eval=F}
form_data %>% select(pets) %>% filter(sports=="Yes")
```

### Cleaning or manipulating a dataset with ``mutate()``

Suppose that you want to make a new variable called ``coffee_drinker`` that stores `TRUE` if a student reports drinking more than 0 cups of coffee per week, and stores `FALSE` otherwise. 

This code will help us do that.

```{r, eval=F}
form_data <- form_data %>% mutate(coffee_drinker = coffee > 0)
```

We interpret this command as follows, starting from the right hand side of the expression: take the dataset called ``form_data``, and change (``mutate``) it to add a new variable, named ``coffee_drinker``. This new variable will store the answer to the question: is ``coffee`` (an existing variable in ``form_data``) greater than $0$? Then, at the end, we take this whole mutated object and save it back to the dataset named ``form_data`` (this is handled by ``form_data <-`` at the start).

Note that, inside of my ``mutate`` statement, I use both ``=`` and ``>``. Here, they serve different purposes. The ``=`` is *assigning* a value to a new variable, called ``coffee``. On the other hand, the ``>`` is *asking a question* to the data. Later, we will see that if we want to *ask a question* about if two things are equal, we need to use ``==`` instead of ``=``. 

You will need to use the ``mutate`` command on HW1, so please ask questions if you are confused!

We can see that our command worked by summarizing this new ``coffee_drinker`` variable.

```{r, eval=F}
form_data %>% select(coffee_drinker) %>% table()
```

I think that this has been enough about extracting, cleaning, and manipulating data to get you started on Homework 1. Let's move on!

# Explore and describe a single quantitative variable

Datasets are usually too big to understand just by looking at the raw data. We need to go one variable at a time and produce numerical and visual summaries. Luckily, R makes this all pretty straightforward. We will start by covering this process for numerical variables.

<div id="boxedtext">

**Goals:**

- Make a histogram; describe what you see.
- Make a boxplot; describe what you see.
- Have R compute measures of center and spread (mean, median, sd, IQR, etc.)

</div>

Let's use ``pets,`` which stores the recorded answers to the question "How many pets have you had or have you lived with in your life?". As a note, since we will not need the ``height`` variable at all today, I am going to work with ``form_data``, not ``form_data_subset``. 

### Make a histogram

Let's first learn how to make a histogram in R. There are actually two ways. The first may look simpler, but I would like to encourage you all to use the second version, which comes from the ``tidyverse`` package.  While at first ``ggplot`` may seem unnecessarily complex, 
the ``tidyverse`` package has become the standard for data cleaning and data visualization. Once you get used to ``ggplot()``, it will let you make very complex and very beautiful, customizable plots. I use ``ggplot()`` for all of my research papers! 

```{r}
hist(form_data$pets)
ggplot(data=form_data, aes(x=pets))+geom_histogram()
```

Play around with the code below. Is there a histogram that you like best?

```{r}
ggplot(data=form_data, aes(x=pets))+geom_histogram(col="red", fill="pink", binwidth=1)+theme_bw()+
  xlab("Pets")+ylab("Frequency")+ggtitle("My favorite histogram")
```

Suppose that, for some reason, you wanted to ignore the student with more than 8 pets. I wouldn't recommend ignoring this student in real data analysis, I just want to make a point about plotting.Try to figure out what the difference is between the two lines of code below:

```{r}
ggplot(data=form_data %>% filter(pets < 8), aes(x=pets))+geom_histogram()+ggtitle("One way to make this histogram")
ggplot(data=form_data, aes(x=pets))+geom_histogram()+ggtitle("Another way to make this histogram")+xlim(0,8)
```
You can always change your axes limits manually in R. By default, R sets your axes limits to capture the entire dataset.

### Make a boxplot

Once again, we have a "base R" version and a ``tidyverse`` (``ggplot``) version. I highly recommend the ``ggplot()`` version: once we start comparing data across groups, it will be super useful.

```{r}
boxplot(form_data$pets)
ggplot(data=form_data, aes(y=pets))+geom_boxplot()
```

**Challenge:** Using what you learned in the histogram section, can you figure out how to add axes labels and a title to your ``ggplot`` boxplot?

*Hint:* It really isn't that challenging, because it looks exactly the same as it did for histograms.

### Report numerical summaries

Everything that we discussed in lecture is easily computed in R. Usually, the name of the relevant ``R`` function is just the english name of the statistic we want to compute: how lucky are we!

The following three lines of code all report the same number. Which do you prefer?

```{r}
mean(form_data$pets)
form_data %>% summarize(mean(pets))
form_data %>% pull(pets) %>% mean()
```

Try running the same lines of code as above, but change the word ``mean()`` to the word ``median()``. What happens? Explain how the mean and the median relate for this dataset.

```{r}
median(form_data$pets)
form_data %>% summarize(median(pets))
form_data %>% pull(pets) %>% median()
```

One of my favorite things about the ``summarize()`` function is that we can report multiple summary statistics at a time. We can also stratify our summaries into groups, which is really helpful. Check out the following lines of code.

```{r}
form_data %>% summarize(mean(pets), median(pets), IQR(pets), sd(pets))
form_data %>% group_by(sports) %>% summarize(mean(pets), median(pets), IQR(pets), sd(pets))
form_data %>% filter(sports=="Yes") %>% summarize(mean(pets), median(pets), IQR(pets), sd(pets))
```

# Explore and describe a single categorical variable

Suppose that we want to explore the variable ``dining``. This is a categorical variable. 

Summarizing a single categorical variable is sort of boring! We don't have measures of center and spread. The main way that we numerically explore a single categorical variable is by making a table. Try out the following lines of code, which all give the same information. What did you learn about this variable?

```{r}
form_data %>% group_by(dining) %>% summarize(n())
table(form_data$dining)
form_data %>% select(dining) %>% table()
```

If we wanted to visually explore a single categorical variable, we would typically use a bar chart. This displays the same information as the table.

```{r}
ggplot(data=form_data, aes(x=dining))+geom_bar()
```

Again, a nice thing about learning ``ggplot()`` is that we can use exactly the same methods as before to make this plot more beautiful- we can just use the plus sign to add on aesthetic features. 

```{r}
ggplot(data=form_data, aes(x=dining, fill="pink"))+geom_bar()+xlab("Favorite Dining Hall")+ylab("Number of Students")+ggtitle("My favorite plot")
```

# The relationship between a numerical variable and a categorical variable

Everything in statistics gets more interesting when we can explore relationships between variables. Suppose that you are interested in knowing if those who play a sport at Williams tend to be taller than those who do not play a sport. 

### Visually exploring the relationship

To start, we need to *clean* the ``height`` variable so that R recognizes it as a number. Be sure that you understand why this is needed in this case!

```{r}
form_data$height
```
Instead of manually fixing people's answers, I will take the lazy route. I will first see if R can automatically convert the heights to numbers, using ``as.numeric()``. 

```{r}
form_data <- form_data %>% mutate(height_numeric = as.numeric(height))
```

If R cannot figure out the correct number on its own, then ``height_numeric`` will be `NA`. We can remove these individuals from our dataset as follows. Note that I do not want to fully remove them from ``form_data``, so I make a new, smaller dataset that removes anyone with a `NA` value for ``height_numeric``.

```{r}
form_data_subset <- form_data %>% filter(!is.na(height_numeric))
```

**Challenge question:** Can you figure out what the `!` does here?

We are now ready to run the following code to begin to make side-by-side boxplots of the ``height_numeric`` variable, separated by categories of the ``sports`` variable. This is our favorite way to visually display the relationship between a numerical and a categorical variable.

```{r}
ggplot(data=form_data_subset, aes(x=sports, y=height_numeric))+geom_boxplot()
```
Be sure that you can describe what you see in this plot.

Side-by-side histograms are a bit less common for displaying the relationship between a categorical and a numerical variable. But there are still a bunch of ways that we could go about making side by side histograms. 

```{r}
ggplot(data=form_data_subset, aes(x=height_numeric))+geom_histogram(binwidth=3)+facet_grid(vars(sports))
```

What do you learn from this plot?

What about trying to use colors to represent ``sports``? This is an option, but it is not my favorite option for histograms. If you have time, you can explore the code below.

```{r}
ggplot(data=form_data_subset, aes(x=height_numeric, fill=sports))+geom_histogram(binwidth=3, position="stack")
ggplot(data=form_data_subset, aes(x=height_numeric, fill=sports))+geom_histogram(binwidth=3, position="dodge")
```


### Numerically summarizing the relationship

We may also want to report numerical summaries of the ``height`` variable across our two groups of the ``sport`` variable. The tidyverse makes this very seamless.

```{r}
form_data_subset %>% group_by(sports) %>% summarize(mean(height_numeric), sd(height_numeric), median(height_numeric))
```

This variable is not particularly skewed, and so the means and the medians are very similar within each group. 

Finally, if you have time, explore these alternate ways to report the same information.

```{r}
form_data_subset %>% filter(height_numeric > 10, sports=="No") %>% summarize(mean(height_numeric), sd(height_numeric), median(height_numeric))
form_data_subset %>% filter(height_numeric > 10, sports=="Yes") %>% summarize(mean(height_numeric), sd(height_numeric), median(height_numeric))
```

## Wrap up

If you were lost in class today, please read through this tutorial again on your own time. Actually run each line of code in the tutorial, and try to think about what it is doing. 

Hopefully, when you work on HW1, you will find that I went over all of the R functions that you need for the homework in class. It is my goal that this will be the case for every homework throughout the semester! 

If you are already experienced in programming, I would encourage you to go above-and-beyond on the homework assignments in terms of making your code and your plots beautiful. If you are new to programming, please let me know if the pacing seems too fast for you! I am happy to provide additional resources for learning R if needed!

Finally, when it comes to debugging, the internet is your friend! Learning to use google, stack overflow, or ChatGPT to resolve R errors on your own is a key skill. Please do not stare at a piece of code for 8 hours waiting for a TA session because you cannot find the source of an error: learn to google your error messages and problem solve for yourself! **Caveat:** always try to write your code on your own first, and ask AI for help *only* with your error messages, not with the entire problem. You need to learn the principles for yourself!

*** 
## Acknowledgements

The formatting and some of the introductory content in this tutorial was adopted from an OpenIntro lab.

<div id="license">
This is a product of OpenIntro that is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel
from a lab written by Mark Hansen of UCLA Statistics.
</div>

