---
title: "Lecture 8: t-tests in R"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(openintro)
```


<div id="boxedtext">

**Learning Objectives**

+ Learn how to use R to test for a difference in population means across two groups. 
+ Reproduce the results that R gives you "by hand" to ensure that you understand where R's results come from. 
+ Get mini-quiz credit for attendance and effort/progress.
</div>

# Loading the data

```{r, echo=F, message=F, warning=F}
library(knitr)
```

We will work with the ``nycflights`` dataset from the ``openintro`` package. Please start by loading the required packages and the dataset. 

```{r, warning=F, message=F, echo=T}
library(tidyverse)
library(openintro)
data(nycflights)
```

Even though this dataset is massive, the ``openintro`` documentation says that this is a random sample of flights from the population of all flights that departed from NYC in 2013. To see the documentation, you can type ``?nycflights`` in your console. 

# Setting up a research question

I would like to know if flights headed to Denver are more delayed, on average, than flights headed to Los Angeles. 

Since my research question relates only to a certain subset of flights in my dataset, my first step is to subset my dataset. 

```{r, eval=T}
nycflights_subsamp <- nycflights %>% 
  filter(dest=="LAX" | dest == "DEN")
```

Next, I can either visually or numerically compare the distribution of flights headed to Los Angeles vs. Denver in my sample. I decided to do this with boxplots, but you could do this however you choose. I came up with the following plot:

```{r, echo=T, fig.height=3, fig.width=2, fig.align="center", echo=F, eval=T}
ggplot(nycflights_subsamp, aes(x=dest, y=dep_delay))+
  geom_boxplot()+ 
  ylab("Departure Delay")+
  xlab("Destination Airport")+theme_bw()
```

Based on the plot, I can tell that there is a difference in means between flights headed to `LAX` and `DEN` in my sample. However, I cannot tell if this is driven by a true underlying population difference, or just a few unlucky outliers in my sample. 

We will use a t-test to get to the bottom of whether or not there is evidence of an underlying population difference here.

# A hypothesis test for a difference in means

Let $\mu_1$ be the population average of delay time for flights headed from NYC to Denver. Let $\mu_2$ be the population average of delay time for flights headed from NYC to Los Angeles. The null hypothesis is
$$
H_0: \mu_1 - \mu_2 = 0.
$$
We want to see if our sample provides evidence that $\mu_1 - \mu_2 \neq 0$. We can actually get R to do all of this work for us. The following line of code will conduct a t-test to see if ``dep_delay`` differs by destination airport in our subsetted dataset.

```{r, eval=T}
t.test(dep_delay~dest, data=nycflights_subsamp)
```

When you run this code, it prints out a lot of information for you! 

At a high level, we see that the p-value is equal to $0.00055$. You should recognize that this means we *reject* the null hypothesis, and conclude that there is convincing evidence of a difference in means. 

But how did `R` come up with this p-value? And what is all of the other information that is printed? We will dive into this below!

# Your task in groups

Your task is to reproduce all of the numbers in the `t.test()` output using the dataset and the theory that we learned in class. 

1. Your sample statistic should have the form $\bar{x}_1 - \bar{x}_2$. Find the values of $\bar{x}_1$ and $\bar{x}_2$ directly from your dataset, and save then in `R` to variables called `xbar1` and `xbar2`. Do these show up in the ``t.test()`` output?

2. R is basing this t-test off of a sampling distribution, whose standard deviation is supposed to be 
$$
\sqrt{\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}}.
$$
However, $\sigma_1$ and $\sigma_2$ are both unknown. Obtain estimates from them using your data, and save the result to variables called `s1` and `s2` in R. While you are at it, also save the appropriate values of $n_1$ and $n_2$ to `R` as `n1` and `n2`. Do any of these numbers appear in the ``t.test()`` output?


3. Finally, plug all of your estimates in to obtain your estimated standard deviation of your sampling distribution:
$$
\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}.
$$
be sure to save this value in `R`, to a variable called ``se``. While this number does not appear in the `t.test()` printout, you should check to see if your answer matches the number printed by the code: ``t.test(dep_delay~dest, data=nycflights_subsamp)$stderr``.

4. The ``t.test()`` output prints out `t=3.46`. Based on what you have calculated so far, reproduce the number 3.46. On your mini quiz, explain briefly what you did.

5. When we do a two-sample t-test for a difference in means and the two samples have different sizes and different standard deviations, a complicated formula tells us how many degrees of freedom our t-test has. In this case, I would argue that we really don't need to worry about the exact degrees of freedom for our test. In the context of this dataset, why might I argue that we don't need to worry about the exact degrees of freedom?

6. Use `R` to reproduce the p-value given by the ``t.test()`` output. You may *either* use the ``pt()`` function and tell it that ``df=1118.2``, or you may simply use ``pnorm()``. You can even try both and compare your answers. On your mini quiz, briefly explain how you computed the p-value. A picture might help. 

```{r}

```


7. Use `R` to reproduce the confidence interval printed by ``t.test()`. Once again, you may use either a `t`-multiplier (with the appropriate degrees of freedom) or a `z`-multiplier. On your mini quiz, explain briefly how you did this, and interpret your confidence interval in your own words. 


